---
title: "Decision Trees - Weather and plane related variables (balanced data)"
author: "Mandip Farmahan"
date: "2023-06-13"
output: html_document
---

# Import statements

## Import libraries

```{r, message=FALSE}
library(tidyverse)
library(here)
library(splitstackshape)
library(janitor)
library(rpart)
library(rpart.plot)
library(ranger)
library(caret)
library(ROSE)
```

## Import dataset

```{r, message=FALSE}
flight_delays_df <- read_csv(here("data/clean_data/flight_delay_clean.csv"))
```

```{r}
ewr_delays_df <- flight_delays_df %>% 
  filter(delay_flag != "Cancelled",
         origin == "EWR") %>% 
  mutate(delay_flag = factor(delay_flag, levels = c("No", "Yes"),
                             labels = c("OnTime", "Delayed")),
         dest = as_factor(dest),
         carrier = as_factor(carrier),
         type = as_factor(type))
```


```{r}
ewr_delays_df %>% 
  count(delay_flag)
```

## Create save files

```{r}
rf_tune_save <- "saved_models/weather_plane_bal_rf_tune.RDS"
```

## Select variables

```{r}
flight_delays_df %>% 
  head(3)
```

```{r}
delay_trim_df <- ewr_delays_df %>%  
  select(delay_flag, wind_dir, wind_speed, pres, tavg, prcp,  
         visib, snow, snwd,
         distance, nose_dir, carrier, type, seats, dest) %>% 
  drop_na(seats)
```

```{r}
summary(delay_trim_df)
```

## Create a train/test split of 80:20

The model will be saved during the first execution and then loaded for subsequent analysis to ensure reproducibility.

```{r}
shuffle_index <- sample(1:nrow(delay_trim_df))
delay_trim_tidy <- delay_trim_df[shuffle_index, ]

test_train_set <- stratified(delay_trim_tidy,
                             group = c("delay_flag", "wind_dir",
                                       "wind_speed"),
                             size = 0.2,
                             bothSets = TRUE)
```

```{r}
delay_test <- test_train_set$SAMP1
delay_train <- test_train_set$SAMP2
```

```{r}
# calculate number of "Delayed" outcomes in train set for rose_train
num_delayed_train <- delay_train %>% 
  summarise(num = sum(delay_flag == "Delayed")) %>% 
  pull()

# create balanced train dataset (Delayed vs OnTime)
rose_train <- ROSE(delay_flag ~ .,
       data  = delay_train,
       N = num_delayed_train * 2)$data
```


```{r}
table(rose_train$delay_flag)
```

```{r}
# calculate weight for rf_classifier (percentage of "Delayed" in train set)
split_weight <- delay_train %>% 
  summarise(perc = sum(delay_flag == "Delayed") / n()) %>% 
  pull()
```

```{r}
# calculate number of "Delayed" outcomes in test set for rose_test
num_delayed_test <- delay_test %>% 
  summarise(num = sum(delay_flag == "Delayed")) %>% 
  pull()

# create balanced test dataset (Delayed vs OnTime)
rose_test <- ROSE(delay_flag ~ .,
                   data  = delay_test,
                   N = num_delayed_test * 2)$data

table(rose_test$delay_flag)
```

## Check test/train set balance

### `delay_flag` alone

```{r}
delay_test %>%
  tabyl(delay_flag)
```

```{r}
delay_train %>%
  tabyl(delay_flag)
```

### `delay_flag` with `wind_dir`

```{r}
delay_test %>%
  tabyl(delay_flag, prcp) %>% 
  adorn_percentages("col")
```

```{r}
delay_train %>%
  tabyl(delay_flag, prcp) %>% 
  adorn_percentages("col")
```


### `delay_flag` with `wind_speed`

```{r}
delay_test %>%
  tabyl(delay_flag, snow) %>% 
  adorn_percentages()
```

```{r}
delay_train %>%
  tabyl(delay_flag, snow) %>% 
  adorn_percentages()
```

The test/train splits appear balanced.

## Build tree model using training dataset

```{r}
delay_fit <- rpart(
  formula = delay_flag ~ ., 
  data = rose_train, 
  method = 'class'
)

rpart.plot(delay_fit,
           yesno = 2,
           faclen = 6,
           digits = 2)
```

```{r}
rf_classifier <- ranger(delay_flag ~ ., 
                        data = delay_train,
                        replace = FALSE,
                        sample.fraction = c(split_weight, split_weight),
                        keep.inbag = TRUE,
                        importance = "impurity", 
                        num.trees = 500, 
                        mtry = 2, 
                        min.node.size = 5)

rf_classifier
```

```{r}
#calculate the number of samples for each outcome used
inbag <- do.call(cbind, rf_classifier$inbag.counts)
# number of OnTime samples
colSums(inbag[delay_train$delay_flag == "OnTime", ])[1]
# number of Delayed samples
colSums(inbag[delay_train$delay_flag == "Delayed", ])[1]
```

```{r}
importance(rf_classifier)
```

```{r}
delay_test_pred <- rose_test %>%
  mutate(pred = predict(rf_classifier,
                        data = rose_test)$predictions)

confusionMatrix(delay_test_pred$pred,
                delay_test_pred$delay_flag)
```

## Random forest tuning

```{r, eval=FALSE}
# set eval=TRUE and comment out readRDS line in next chunk to run tuning block
control <- trainControl(
  method = "repeatedcv", 
  number = 5, 
  repeats = 10
)

tune_grid = expand.grid(
  mtry = 1:6,
  splitrule = c("gini", "extratrees"),
  min.node.size = c(1, 3, 5)
)

rf_tune <- train(
  delay_flag ~ .,
  data = rose_train,
  method = "ranger",
  metric = "Kappa",
  num.trees = 50,
  importance = "impurity",
  tuneGrid = tune_grid, 
  trControl = control
)
```


```{r}
rf_tune <- readRDS(rf_tune_save)

plot(rf_tune)
```

```{r}
rf_tuned_classifier <- ranger(delay_flag ~ .,
                              data = delay_train,
                              replace = FALSE,
                              sample.fraction = c(split_weight, split_weight),
                              keep.inbag = TRUE,
                              importance = "impurity",
                              num.trees = 500,
                              mtry = 6,
                              min.node.size = 5,
                              splitrule = "gini")
```

```{r}
importance(rf_tuned_classifier)
```

```{r}
#calculate the number of samples for each outcome used
inbag <- do.call(cbind, rf_tuned_classifier$inbag.counts)
# number of OnTime samples
colSums(inbag[delay_train$delay_flag == "OnTime", ])[1]
# number of Delayed samples
colSums(inbag[delay_train$delay_flag == "Delayed", ])[1]
```



```{r}
delay_tuned_test_pred <- rose_test %>%
  mutate(pred = predict(rf_tuned_classifier,
                        data = rose_test)$predictions)

confusionMatrix(data = delay_tuned_test_pred$pred,
                reference = delay_tuned_test_pred$delay_flag)
```

```{r}
# untuned model for comparison 
confusionMatrix(delay_test_pred$pred,
                delay_test_pred$delay_flag)
```

